### 2023-12-05 thesis/MSECE_thesis.tex
Grammar revisions.

\section{Abstract}\label{sec:abstract}
With recent advances in machine learning and deep learning technologies along with the creation of larger aviation-specific corpora, the application
of natural language processing technologies to aviation communications, especially those based on transformer neural networks, is becoming
increasingly feasible. Previous work has been done focusing on machine learning applications to natural language processing such as N-grams and word
lattices. This thesis will layout a process for pretraining transformer-based language models on aviation English corpora and compare the
effectiveness and performance of language models transfer learned from pretrained checkpoints and those trained from their base weight initializations
(trained from scratch). The results show that transformer language models trained from scratch outperform models fine-tuned from pretrained
checkpoints, achieving perplexity values on the test partition of the corpus nearly two orders of magnitude lower than that of the pretrained models.
The work concludes by making recommendations for future work to improve pretraining performance as well as suggestions for downstream, in-domain tasks
such as semantic extraction, named entity recognition (callsign identification), speaker role identification and speech recognition.

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural language processing
and language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
be brought to consumer facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general
domain, applications have begun to spin-off into problem-specific domains, such as (and most notably for this work) aviation. For example, several
speech recognition systems with rule- and machine learning-based natural language processing, inference, and semantic extraction algorithms have been
implemented at EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI augmented air traffic
management (ATM) systems on controller workloads. Preliminary studies have found that AI augmented ATM systems not only reduce controller workloads
but also reduce fuel consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times thereby reducing fuel
consumption~\cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the availability of large
natural language corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain, the
aviation domain still prefers, and sees much success with, traditional machine learning models for language modeling applications. Typically, the most
prevalent of the traditional models are N-grams, word lattices, and/or rule-based approaches depending on the application and operating environment of
the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the
aviation domain. By comparison, the largest corpora in the general domain is either WikiText or the BookCorpus which contain trillions of tokens
whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to
transcribe/label data as it requires transcribers to have domain knowledge and experience in order to create effective transcriptions. This, in
addition to the following facets, further contribute to the difficulty of transcription: (1) the medium of communication, (2) the environment in which
the communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive
    communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4)
an active Air Traffic Control (ATC) tower and/or channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.

% connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question
Answering Dataset (SQuAD) benchmark has recorded results consistently better than human since
2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
aviation domain. If they can achieve better-than-human performance in the aviation domain, and be used in conjunction with pilot and air traffic
controller systems, they could easily boost efficiency, reduce workloads, and reduce and/or mitigate errors during normal operations.

The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep learning
based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the language
models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the results show
that language models trained ``from scratch'' on in-domain data perform better than model trained on out-of-domain data and transfer learned on
in-domain data i.e.~pretrained BERT and RoBERTa models transfer learned to aviation English.

\section{Notation and Terminology}
This section is included for the sake of clarity throughout this thesis. In pursuit of that, there are a few terms and notations that are defined
below.

Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field
    of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms,
to convey mathematical concepts. With that being said, the shorthand below is defined
\begin{equation*}
    A; x := A \cup \{x\}
\end{equation*}

@@

\section{Literature Review}\label{sec:lit_review}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
state-of-the-art \cite{vaswani_attention_2017}. Transformer architectures have been found to be extremely effective at learning representations and
understandings of languages to predict token probabilities as opposed to transforming one language into another
\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP-related tasks such as prompt completion
and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively)
\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019}proposed pretraining methods such as masked language modeling and
next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et
al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.

@@

% LMs and NLP in aero. domain
Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications
and try to mitigate safety incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict
contributing factors and topic models\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing
has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and
heading info) into an XML language structure\cite{ragnarsdottir_language_2003}. Transformer language models such as BERT and RoBERTa have been applied
to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks, translation tasks (between notations; e.g.~NOTAM notation to
Airlang to make parsing tasks easier) and reduce the workload for pilots during long-haul flights\cite{arnold_knowledge_2022}. Transformer models
have also been used for speaker role identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications),
specifically, a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed well at speaker
and role identification tasks in general\cite{guo_comparative_2022}.

% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification,
etc. A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language
modeling is used to transcribe speech
\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
and word lattices have also been used in the aviation domain alongside semi-supervised approaches to automatic speech recognition to leverage
unlabeled data. However, none of the developed methodologies use neural language models for their applications
\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.

% differences in language (transcripts vs literary english), narrow model selection
The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation
English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the
similarity (or, conversely, the difference) between literary English and aviation English. If the two domains are sufficiently different, then
training the pretrained models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text
domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models
\cite{raffel_exploring_2020}. Yadlowsky et al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different,
model pretraining on data considered out-of-domain will have negative effects on model performance.

\section{Data Sources}\label{sec:data_source}
There are four corpora used in this thesis. The datasets are primarily speech corpora intended either for automatic speech recognition research in the
domain of aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and
transcriptions, which are audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers
(described in sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).

The four corpora that were selected and used for this work are briefly described below. Further analysis is performed in Section
\ref{sec:data_analysis}.

\subsection{Air Traffic Control Complete}\label{sec:atcc}
The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions.
ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and
transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded through the placement of Very
High Frequency (VHF) antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types
of frequencies monitored vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora
\cite{godfrey_air_1994}. See Table \ref{tab:atcc_stats} for corpus-specific statistics.

@@

\subsection{ATCO2}\label{sec:atco2}
The ATCO2 dataset is a speech corpus that consists of audio communications at Prague and Brno airports, in Czechia, along with corresponding
transcriptions. The speech recordings and transcripts are crowd sourced from volunteers. The dataset was created and labeled, in part, as an English
language detection corpus, however, the labeling also includes transcripts of speech segments which makes it practical for ASR and language model
development in addition to language detection (although in this work it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
See Table \ref{tab:atco2_stats} for corpus-specific statistics.

@@

\subsection{Air Traffic Control Simulation}\label{sec:atcosim}
The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions.
The data was recorded at the EUROCONTROL Experimental Centre for air traffic control simulation in Bretigny-sur-Orge, France. In this corpus, only the
controllers' voices are included and therefore the transcripts only include the controller's side of each interaction. The audio data was transcribed
by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was transcribed, it was reviewed and
corrected and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}. See Table
\ref{tab:atcosim_stats} for corpus-specific statistics.

@@

\subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both controller and pilot sides of the
communications are included in this data. Annotations were created by experienced transcribers/annotators and labeled samples were randomly selected
for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and
unified\cite{smidl_air_2019}. See Table \ref{tab:zcu_cz_atc_stats} for corpus-specific statistics.

@@

\section{Data Processing}\label{sec:data_processing}
All of the corpora listed above (see Section \ref{sec:data_source}) are created at different times, for different purposes, and by different authors.
As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal
being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).

All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used
    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by parsing from the
corpora-specific formats where necessary i.e.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations
are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text
is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus).

\subsection{Handling Spelling Errors \& Inconsistencies}\label{sec:spelling_errors}
By the nature of the data labeling process, there is a high likelihood to be spelling errors introduced by the annotators during corpora creation.
This gets exacerbated by the aggregation of several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of
the errors are recoverable/correctable given context. For example, ``possibility of \textbf{tornado's}'' should clearly have been ``possibility of
\textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions that are
consistent within corpora, but become inconsistent when the corpora are combined.

% 22 spelling errors
Manually reviewing every sample in the corpus is unrealistic considering that there are over 50,000 data samples and would be subject the same
human error that introduced those errors in the first place. Instead, the frequency of occurrence of tokens in the corpora combined with manual review
thereafter, was used to detect and correct the most common errors. The frequency of occurrence of a token is calculated using equation
(\ref{eq:token_frequency_occurrence}).

@@

\noindent
Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the
total number of tokens in the corpus. Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and
corrected where necessary. An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the
corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed, and if an erroneous token, $A$, is found, it is replaced
with the corrected version, $B$. A total of 22 unique errors were detected and subsequently corrected or removed using this method, although multiple
instances of those errors typically occurred.

% hesitations, cut off words, different kinds of filled pauses
The mapping system described above is also used to modify within corpora conventions to be consistent across corpora. Three areas are addressed here:
hesitations, cut-off or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes
unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as any incomplete or non-english speech sounds and left
completely undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are
transcribed differs between corpora as well. Two corpora prescribe special tokens to these phenomena such as ``[hes]'', ``[hesitation]'', ``<pause>'',
or similar, while the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for
instances that were cut-off mid pronunciation). In order to facilitate the unification of the transcripts for all four corpora, the conventions for
transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.

@@

A special token is used to represent hesitations when no other context regarding the realization of hesitation is present. If other corpora use a
special token to represent hesitations, it is translated to match the convention in Table \ref{tab:phenomena_conventions}. Partial/cut-off words
are represented with a hyphen (-) at the point where the pronounced word is cut off. For example, if the last syllable of ``approaching'' is cut
mid transmission, it would be represented in the transcript as ``approa-'' or, likewise, if the first syllable is cut off, it would be written as
``-roaching''. Lastly, if there is enough information in the original transcript to indicate the realization of a hesitation, it is represented by
the most appropriate token from ``Filled pauses'' in Table \ref{tab:phenomena_conventions}.

One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts,
so including as much data as possible regarding the realization of spoken words is useful given the context. The \lstinline|[HES]| token is used as a
catch-all, fallback token in the case that no knowledge can be extracted about the utterance from the transcript.

\section{Data Analysis}\label{sec:data_analysis}
% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in
the corpus, mean sequence length per sample in the corpus, the number of unique tokens in the corpus, and the total number of tokens in the corpus.
The same statistics were also calculated across corpora, since all four corpora will be combined for training and testing the language and speech
recognition models. Note that this is based on the raw data after loading and unifying the corpora, prior to preprocessing for model training/testing
and outlier removal.

@@

The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined
data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data. The sequence lengths of the samples are fairly consistent across
corpora with at most about an 18\% percent difference between sequence lengths (calculated using equation (\ref{eq:percent_difference}), below). The
number of unique tokens does not scale additively to the combined corpora, which is as expected since there will be an overlap of tokens between
corpora. Aviation English generally conforms to and follows ICAO phraseology standards, so we would expect that many terms would be
shared across corpora with the exception of geographic/location specific terms, such as city, landmark, airport, and airline names, etc.~that
typically vary across regions.

@@

\noindent
Except for the first token in the sequence, the probability of a token occurring at a specific point in the sequence depends on the tokens preceding it
\footnote{The most common use case for an N-gram model is the preceding tokens.}. Due to this fact, N-grams are sometimes referred to as
left-associative models (in a left-to-right language).

>>>>

\section{Abstract}\label{sec:abstract}
With recent advances in machine learning and deep learning technologies and the creation of larger aviation-specific corpora, applying
natural language processing technologies to aviation communications, especially those based on transformer neural networks, is becoming
increasingly feasible. Previous work has focused on machine learning applications to natural language processing, such as N-grams and
word lattices. This thesis will lay out a process for pretraining transformer-based language models on aviation English corpora and
compare the effectiveness and performance of language models transfer learned from pretrained checkpoints and those trained from their
base weight initializations (trained from scratch). The results show that transformer language models trained from scratch outperform
models fine-tuned from pretrained checkpoints. The work concludes by recommending future work to improve pretraining performance and
suggestions for downstream, in-domain tasks such as semantic extraction, named entity recognition (callsign identification), speaker
role identification, and speech recognition.

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural
language processing and language modeling technologies, in particular, have made tremendous advances over the past two decades,
achieving performance levels good enough to be brought to consumer-facing products such as personal assistants, search engines, phones,
televisions, etc. Due to the success in the general domain, applications have begun to spin off into problem-specific domains, such as
(and most notably for this work) aviation. For example, several speech recognition systems with rule- and machine learning-based
natural language processing, inference, and semantic extraction algorithms have been implemented at EUROCONTROL simulation centers in
Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM) systems on
controller workloads. Preliminary studies have found that AI-augmented ATM systems reduce controller workloads and fuel consumption of
aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times, thereby reducing fuel consumption
\cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a variety of deep learning methods/architectures, namely, recurrent, convolutional, and transformer neural networks, due to the availability of large natural language corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain, the aviation domain still prefers and sees much success with traditional machine learning models for language modeling applications. Typically, the most prevalent of the traditional models are N-grams, word lattices, and rule-based approaches, depending on the application and operating environment of the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the aviation domain. By comparison, the largest corpora in the general domain are either WikiText or the BookCorpus, which contain trillions of tokens, whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000 tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to transcribe/label data as it requires transcribers to have domain knowledge and experience to create effective transcriptions. This, in addition to the following facets, further contributes to the difficulty of transcription: (1) the medium of communication, (2) the environment in which the communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.

% connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question Answering Dataset (SQuAD) benchmark has recorded results consistently better than humans since 2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/})~\cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the aviation domain. If they can achieve better-than-human performance in the aviation domain and be used with pilot and air traffic controller systems, they could easily boost efficiency, reduce workloads, and reduce and mitigate errors during normal operations.

The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep learning-based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the results show that language models trained ``from scratch'' on in-domain data perform better than models trained on out-of-domain data and transfer learned on in-domain data, i.e., constrained BERT and RoBERTa models transfer learned to aviation English.

\section{Notation and Terminology}
This section is included for the sake of clarity throughout this thesis. In pursuit of that, a few terms and notations are defined below.

Set notation (or set-builder notation)\footnote{That is, the notation used to denote sets and associated operations therein as it relates to the field of set theory, in mathematics.} as defined in \cite{jech_chapter_1978} is used in parts of this thesis, particularly in equations and algorithms, to convey mathematical concepts. The shorthand below is defined
\begin{equation*}
    A; x := A \cup \{x\}
\end{equation*}

@@

\section{Literature Review}\label{sec:lit_review}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art \cite{vaswani_attention_2017}. Transformer architectures are extremely effective at learning representations and understandings of languages to predict token probabilities instead of transforming one language into another \cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively) \cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019}proposed pretraining methods such as masked language modeling and next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.

@@

% LMs and NLP in aero. domain
Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications and try to mitigate safety incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}. Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict contributing factors and topic models\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and heading info) into an XML language structure\cite{ragnarsdottir_language_2003}. Transformer language models such as BERT and RoBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks, translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload for pilots during long-haul flights\cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications); specifically, a pretrained BERT model was used and fine-tuned on problem-specific data and compared to other models that performed well at speaker and role identification tasks in general\cite{guo_comparative_2022}.

% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification, etc. A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language modeling is used to transcribe speech \cite{guo_comparative_2022,smidl_air_2019,zuluaga gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}. Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams and word lattices, have also been used in aviation alongside semi-supervised approaches to automatic speech recognition to leverage unlabeled data. However, none of the developed methodologies use neural language models for their applications \cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi supervised_2017,badrinath_automatic_2022}.

% differences in language (transcripts vs literary english), narrow model selection
The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the similarity (or, conversely, the difference) between literary English and aviation English. If the two domains are sufficiently different, then training the pretrained models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models \cite{raffel_exploring_2020}. Yadlowsky et al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different, model pretraining on out-of-domain data will negatively affect model performance.

\section{Data Sources}\label{sec:data_source}
Four corpora are used in this thesis. The datasets are primarily speech corpora intended for automatic speech recognition research in aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and transcriptions, which are audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).

The four corpora selected and used for this work are briefly described below. Further analysis is performed in Section
\ref{sec:data_analysis}.

\subsection{Air Traffic Control Complete}\label{sec:atcc}
The Air Traffic Control Complete (ATCC) corpus is a speech corpus consisting of audio recordings with corresponding transcriptions. ATCC is a collection of three smaller corpora recorded at Dallas-Fort Worth, Logan International, and Washington National airports and transcribed by current or former air traffic controllers familiar with the respective areas. Audio data was recorded by placing Very High Frequency (VHF) antennae configured to monitor several frequencies at each airport, such as arrival, departure, approach, ground, etc. The types of frequencies observed vary by airport, but the presence of departure, approach, and ground frequencies is relatively consistent across subcorpora \cite{godfrey_air_1994}. See Table \ref{tab:atcc_stats} for corpus-specific statistics.

@@

\subsection{ATCO2}\label{sec:atco2}
The ATCO2 dataset is a speech corpus comprising audio communications at Prague and Brno airports in Czechia and corresponding transcriptions. The speech recordings and transcripts are crowd-sourced from volunteers. The dataset was created and labeled, in part, as an English language detection corpus. However, the labeling also includes transcripts of speech segments, which makes it practical for ASR and language model development in addition to language detection (although in this work, it is only used for ASR and LM development)~\cite{szoke_detecting_2021}.
See Table \ref{tab:atco2_stats} for corpus-specific statistics.

@@

\subsection{Air Traffic Control Simulation}\label{sec:atcosim}
The Air Traffic Control Simulation (ATCOSIM) corpus is a speech corpus consisting of audio recordings and corresponding transcriptions. The data was recorded at the EUROCONTROL Experimental Centre for air traffic control simulation in Bretigny-sur-Orge, France. In this corpus, only the controllers' voices are included, and therefore, the transcripts only include the controller's side of each interaction. The audio data was transcribed by one person, trained according to the guidelines and formatting requirements of the corpus. After all data was transcribed, it was reviewed and corrected, and any remaining problems were reviewed by an operational air traffic controller and resolved \cite{hofbauer_atcosim_2008}. See Table \ref{tab:atcosim_stats} for corpus-specific statistics.

@@

\subsection{ZCU CZ ATC Corpus}\label{sec:zcu_atc}
The ZCU CZ ATC corpus consists of audio recordings and corresponding transcripts in the Czech airspace. Both the controller and pilot sides of the communications are included in this data. Experienced transcribers/annotators created annotations, and labeled samples were randomly selected for review during the annotation process. After the dataset was completely labeled, all samples were checked, corrected, and unified\cite{smidl_air_2019}. See Table \ref{tab:zcu_cz_atc_stats} for corpus-specific statistics.

@@

\section{Data Processing}\label{sec:data_processing}
All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors. As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).

All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used, although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by parsing from the corpora-specific formats where necessary, e.g.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus).

\subsection{Handling Spelling Errors \& Inconsistencies}\label{sec:spelling_errors}
By the nature of the data labeling process, there is a high likelihood of spelling errors introduced by the annotators during corpora creation. This gets exacerbated by aggregating several corpora with varying controls for correcting spelling errors and inconsistencies. Nearly all of the mistakes are recoverable/correctable, given proper context. For example, ``possibility of \textbf{tornado's}'' should have been ``possibility of \textbf{tornadoes}'', changing the possessive form of ``tornado'' to the plural form. There are also instances of spelling conventions that are consistent within corpora but become inconsistent when combined.

% 22 spelling errors
Manually reviewing every sample in the corpus is unrealistic, considering that there are over 50,000 data samples and would be subject to the same human error that introduced those errors in the first place. Instead, the frequency of token occurrence in the corpora, combined with manual review thereafter, was used to detect and correct the most common errors. The frequency of occurrence of a token is calculated using equation (\ref{eq:token_frequency_occurrence}).

@@

\noindent
Where $f_w$ is the frequency at which a token occurs in a corpus, $\phi_w(w)$ is the number of times a token, $w$, occurs, and $D_{tokens}$ is the total number of tokens in the corpus. Tokens with the lowest frequency across corpora (and lowest number of occurrences) were manually reviewed and corrected where necessary. An $A \rightarrow B$ mapping system was used to correct the errors, where \(A\) is the erroneous token and \(B\) is the corrected or replaced token. As the text is parsed from the corpora, each sample is analyzed. If an erroneous token, $A$, is found, it is replaced with the corrected version, $B$. 22 unique errors were detected and subsequently corrected or removed using this method. However, multiple instances of those errors typically occurred.

% hesitations, cut off words, different kinds of filled pauses
The mapping system described above is also used to modify corpora conventions to be consistent across corpora. Three areas are addressed here: hesitations, cut-off, or partial words, and filled pauses within speech. These areas are defined differently between the four corpora, which makes unifying the transcripts difficult. Hesitations, for example, are defined by one corpus as incomplete or non-English speech sounds and left completely undefined by another (despite annotators making notes about hesitations). The way in which hesitations, pauses, and partial words are transcribed differs between corpora as well. Two corpora prescribe special tokens to these phenomena, such as ``[hes]'', ``[hesitation]'', ``<pause>'', or similar. In contrast, the others transcribe the data as it sounds (i.e., ``uh'', ``eh'', or ``er'' for filled pauses and ``cir-'' or ``cir+'' for instances that were cut-off mid pronunciation). To facilitate the unification of the transcripts for all four corpora, the conventions for transcribing these phenomena are redefined in Table \ref{tab:phenomena_definitions}.

@@

A special token represents hesitation when no other context is present regarding the realization of hesitation. If other corpora use a special token to express hesitations, it is translated to match the convention in Table \ref{tab:phenomena_conventions}. Partial/cut-off words are represented with a hyphen (-) at the point where the pronounced word is cut off. For example, if the last syllable of ``approaching'' is cut mid-transmission, it would be represented in the transcript as ``approa-'' or, likewise, if the first syllable is cut off, it would be written as ``-roaching''. Lastly, if there is enough information in the original transcript to indicate the realization of a hesitation, it is represented by the most appropriate token from ``Filled pauses'' in Table \ref{tab:phenomena_conventions}.

One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts, so including as much data as possible regarding spoken word realization is useful given the context. The \lstinline|[HES]| token is used as a catch-all fallback token if no knowledge about the utterance from the transcript can be extracted.
\section{Data Analysis}\label{sec:data_analysis}
% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in the corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the corpus. The same statistics were also calculated across corpora since all four corpora will be combined for training and testing the language and speech recognition models. Note that this is based on the raw data after loading and unifying the corpora before preprocessing for model training/testing and outlier removal.

@@

The ATCC and ZCU CZ ATC corpora make up a significant portion of the data, ATCC alone makes up more than half (approximately 54\%) of the combined data, and ATCC and ZCU CZ ATC together make up approximately 81\% of the data. The sequence lengths of the samples are fairly consistent across corpora with at most about an 18\% percent difference between sequence lengths (calculated using equation (\ref{eq:percent_difference}), below). The number of unique tokens does not scale additively to the combined corpora, which is expected since there will be an overlap of tokens. Aviation English generally conforms to and follows ICAO phraseology standards, so we expect many terms to be shared across corpora except geographic and location-specific terms, such as city, landmark, airport, and airline names, which vary across regions.

@@

\noindent
Except for the first token in the sequence, the probability of a token occurring at a specific point depends on the tokens preceding it\footnote{The most common use case for an N-gram model is the preceding tokens.}. Due to this fact, N-grams are sometimes called left-associative models (in a left-to-right language).
